A Survey of Quantitative Descriptions of Molecular Structure

### 2012-07-03 paper.tex
some minor modifications, ready for submission

\maketitle
\begin{abstract}
  Numerical characterization of molecular structure is a first step in
  every computational analysis of molecular data. These numerical
  representations, termed descriptors, come in many forms, ranging
  from simple atom counts and invariants of the molecular graph to
  distribution of properties, such as charge, across a molecular
  surface. In this article we first present a broad categorization of
  descriptors and then describe applications and toolkits that can be
  employed to evaluate them. We highlight a number of issues
  surrounding molecular descriptor calculations such as versioning and
  reproducibility and describe how some toolkits have attempted to
  address these problems.
\end{abstract}

@@

Computational methods play an important role in many chemical
disciplines ranging from drug discovery to materials science. While
there are a variety of techniques that differ in terms of
computational complexity, time requirements and so on. However the
common requirement underlying all these methods is a formal
description of a the molecular structure. Of course, there are many
ways to ``describe'' a molecule. A common approach is to describe the
connectivity, taking into account the types of atoms and bonds. In
other words, explicit representations of chemical structure such as
SMILES, MDL/Symyx SD files and so on. While these descriptions are vital to
modern chemical information systems, they do not necessarily allow
computational techniques to be directly applied to them.

Methods that aim to predict chemical and biological properties
generally require a numerical description of chemical structures. Such
numerical forms range from a set of 3D coordinates (which coupled with
appropriate atom types, is sufficient for methods such quantum mechenical (QM) approaches
and docking) to more abstract numerical descriptions derived from 2D
or 3D representations which can be useful is statistical
approaches. It is now possible to evaluate thousands of numerical
descriptors of chemical structure. As will be discussed later, many of
these descriptors are closely related or capture the same information,
causing that one can be substituted for another. The choice of descriptors is a well-known problem and given a
large collection of them, approaches to identify a suitable subset
have been discussed extensively in the literature~\cite{Miller:2002aa,Kohavi:1997gf}.

In addition to there being many possible descriptors defined in the
literature, there are also multiple implementations of a give
descriptor. These implementations are available in the form of
libraries (which require one to write a program to use them) or
complete applications (graphical user interface (GUI) or command line). As a result, not only
must one choose one or more descriptors that are relevant to the
problem at hand, but one must be concerned about how they
are calculated and whether such a calculation can be reproduced across
different implementations of those descriptors. It is easy to
understand \emph{why} two implementations of the same descriptor can
lead to different results - the primary reasons being that the chemistry
model that underlies the framework or toolkit used to implement the
descriptor is different, and that used data such as isotopic masses
are different.

For example, a descriptor that calculates the number of
aromatic atoms may be implemented using two toolkits with differing
aromaticity models, and hence it is possible that the values generated
by the two implementations will differ. Other sources of differences
include parameters that may be involved in the descriptor calculation
and reference data values (such as atomic radii, electronegativity
values) that are employed during descriptor calculation. While most
implementations will employ the same data sources for standard
concepts (e.g., atomic weights), slight differences in these types of
input data can lead to differences in the final descriptor value~\cite{Guha:2006ac}. As a
result, in most cases, two implementations of a descriptor do not
usually give the exact same value, though they are usually quite
similar. Explicitly explaining the differences may or may not be
possible (it is usually more difficult in the cases of commercial
implementations for which source code is not available)~\cite{Spjuth2010}.

@@

We have noted that a molecular structure can be characterized using a
number of numerical descriptors. In this section we describe a
categorization of descriptors. Admittedly, the grouping is somewhat
arbitrary but serves as a useful guide. In addition, the
categorization also considers the nature of the chemical structure
being considered - some descriptors are only useful when applied to
small molecules, whereas other descriptors are defined specifically
for polymers or protein structures.
Broadly, one can classify descriptors based on the nature of
structural information they require.

@@

Topological descriptors take into account connectivity along with atom
and bond labels. These descriptors essentially consider the molecule
as a labeled graph and characterize it using graph invariants. In
addition to graph invariants, topological descriptors based on
information theory such as entropy \cite{Dehmer:2009uq}. An advantage
of these descriptors is that do not require any intensive
pre-processing steps such as 3D coordinate generation and
conformational analysis. In addition, many graph invariants can be
rapidly computed for graphs corresponding to small molecule structures
(even though they may be computationally intractable on much larger
graphs). While this class of descriptors have been used extensively in
predictive modeling applications
\cite{Garcia-Domenech:2008aa,Randic:2001ad,Besalu:2001aa},
a key drawback is their lack of interpretability. Given the abstract
nature of many graph invariants and information theoretic descriptors,
their use in predictive models makes it difficult to explain the model
in simple physicochemical terms, though various attempts have been  
made at interpretations \cite{Todeschini:1975dq,Stanton:2003aa}.

Geometric descriptors require a 3D conformation as input and therefore
are not as fast as topological descriptors which do not require the 3D
coordinate generation step. Nearly all geometric descriptors work on a
single conformation and for cases where multiple conformation must be
considered, an averaging procedure is usually employed. Examples of
these descriptors include the gravitational index
\cite{Katritzky:1996ly} and moment of inertia descriptors. However,
geometric descriptors encompass many aspects of a molecular
structure. For example, a number of geometric descriptors characterize
the molecular surface. The simplest such example is the surface area
descriptor (and the corresponding volume descriptor). A simple
extension of such a descriptor is to characterize distributions of a
physicochemical property over the molecular surface such as the
Charged Partial Surface Area descriptors \cite{Stanton:1990aa} that
characterize the partial charge distribution over the molecular
surfce. Note that surface descriptor calculations based on analytical
\cite{Connolly:1983aa} or tesselation algorithms
\cite{Eisenhaber:1995qf} can be slow and that parametrized methods
such as the Topological Polar Surface Area (TPSA) \cite{Ertl:2000aa}
and VSA descriptors \cite{Labute:2008aa} that are based on
precalculated surface area values derived from a list of functional
groups.

It is important to note that there is a difference between the information
a descriptor captures and how it is calculated. The above categories are
based on input requirements. This gives rise to situations where
geometrical molecular volume property can be described by geometrical descriptors,
e.g. using an analytical approaches ~\cite{Eisenhaber:1995qf}, and
with approximate, constitutional descriptors, such as the very fast algorithm
use by the VABC approach~\cite{Zhao:2003fk} which is based on group contributions.
The latter descriptor has been
implemented in the CDK and correlates well with other approximations
to the molecular volume (Figure \ref{fig:molvol}).

@@

\subsection{Describing other chemistry}

@@

\textbf{Descriptors for polymeric systems} Polymeric systems represent
a challenge for descriptor development. There have been reports that
employed traditional small-molecule descriptors
\cite{Mattioni:2002dq,Katritzky:1998cr} that consider the polymer in
terms of the individual monomers. A number of approaches have been
described that includes both deterministic, geometry derived
descriptors \cite{Edvinsson:2003bh} as well as stochastic methods
\cite{Gonzales-Diaz:2003ly}. For the case of polypeptides the
descriptors generally tend to be based on the individual amino
acids. Examples include the VHSE family of descriptors described by
Mei et al \cite{Mei:2005kx}, a set of side chain descriptors from
Collantes and Dunn \cite{Collantes:1995vn}, field based descriptors
\cite{Norinder:1991ys} and the AAindex database
\cite{Kawashima:1999zr} which allows rapid lookup of amino acid
properties.

@@

\textbf{Crystal structures} Other methods developed for crystal structures
have been developed too, for example, using a radial distribution
functions~\cite{Willighagen2005}, which capture the packing of organic
crystal structures. Such approaches have been used to cluster
crystal structures~\cite{Wehrens2006} and compare properties~\cite{Willighagen2007}.
% FIXME: add recent work

@@

However, there are certain scenarios where some of the above
characteristics are \emph{undesirable}. One such scenario is in the
safe exchange of structures, such that two parties can perform
analyses on the structure-derived data, without having to know the
structures explicitly. In such cases, the ability to go from a set of
descriptor values back to a chemical structure should be minimized. It
has not been conclusively proven that such a ``one-way'' descriptor
can be defined. For example Masek et al \cite{Masek:2008aa} have shown
that one can regenerate a target structure (or substructural features)
or a set of closely related analogs given a combination of descriptors
such as BCUTs. Tetko \textit{et al.} \cite{Tetko:2005fk} suggest an approach to
prevent reverse engineering but avoid the use of descriptors and
instead sharing structures of related molecules, rather than the
molecules of interest. One could also employ degenerate descriptors,
which would represent a large family of molecules, such that the
target molecule would be hidden. This is likely unreliable as it would
be difficult to perform analyses (e.g., predtive models) with such
descriptors.

@@

As noted above, different implementations of the same descriptor can
generate slightly different values depending on the underlying
chemistry model, differences in reference values and so on. Even
within the same software implementation, different versions of the
same descriptor can generate different values. Thus to maintain
reproducibility in studies that employ moleculad2a28dc0c273a4e060c2fea503adc36er descriptors, some form
of versioning is crucial. The simplest form of version is to manually
note the version of the implementation used for the calculation and
report it along with the results. Though useful, it is not amenable to
automation. In some cases, toolkits will provide an API call to
retrieve the version information, which can then be associated with a
descriptor calculation.

>>>>

\maketitle
\begin{abstract}
  Numerical characterization of molecular structure is a first step in
  many computational analysis of chemical structure data. These numerical
  representations, termed descriptors, come in many forms, ranging
  from simple atom counts and invariants of the molecular graph to
  distribution of properties, such as charge, across a molecular
  surface. In this article we first present a broad categorization of
  descriptors and then describe applications and toolkits that can be
  employed to evaluate them. We highlight a number of issues
  surrounding molecular descriptor calculations such as versioning and
  reproducibility and describe how some toolkits have attempted to
  address these problems.
\end{abstract}

@@

Computational methods play an important role in many chemical
disciplines ranging from drug discovery to materials science.
There are a plethora of techniques that differ in terms of
computational complexity, time requirements and so on. However the
common requirement underlying all these methods is a formal
description of a the molecular structure. There are many
ways to ``describe'' a molecule. A common approach is to describe the
connectivity, taking into account the types of atoms and bonds. In
other words, explicit representations of chemical structure such as
SMILES, MDL/Symyx SD files and so on. While these descriptions are vital to
modern chemical information systems, they do not necessarily allow
computational techniques to be directly applied to them.

Methods that aim to predict chemical and biological properties
generally require a numerical description of chemical structures. Such
numerical forms range from a set of 3D coordinates (which coupled with
appropriate atom types, is sufficient for methods such as quantum
mechanical (QM) approaches and docking to more abstract numerical
descriptions derived from 2D or 3D representations which can be useful
in statistical approaches. It is now possible to evaluate thousands of
numerical descriptors of chemical structure. As will be discussed
later, many of these descriptors are closely related or capture the
same information, allowing one to be substituted for another. The
choice of descriptors is a well-known problem and given a large
collection of them, approaches to identify a suitable subset have been
discussed extensively in the
literature~\cite{Miller:2002aa,Kohavi:1997gf}.

In addition to there being many possible descriptors defined in the
literature, there are also multiple implementations of a give
descriptor. These implementations are available in the form of
libraries (which require one to write a program to use them) or
complete applications (graphical user interface (GUI) or command
line). As a result, not only must one choose one or more descriptors
that are relevant to the problem at hand, but one must be concerned
about how they are calculated and whether such a calculation can be
reproduced across different implementations of those descriptors. It
is easy to understand \emph{why} two implementations of the same
descriptor can lead to different results - the primary reasons being
that the chemistry model that underlies the framework or toolkit used
to implement the descriptor is different. For example, a descriptor
that calculates the number of aromatic atoms may be implemented using
two toolkits with differing aromaticity models, and hence it is
possible that the values generated by the two implementations will
differ. Other sources of differences include parameters that may be
involved in the descriptor calculation and reference data values (such
as atomic radii, electronegativity values) that are employed during
descriptor calculation. While most implementations will employ the
same data sources for standard concepts (e.g., atomic weights), slight
differences in these types of input data can lead to differences in
the final descriptor value~\cite{Guha:2006ac}. As a result, in most
cases, two implementations of a descriptor do not usually give the
exact same value, though they are usually quite similar. Explicitly
explaining the differences may or may not be possible (it is usually
more difficult in the cases of commercial implementations for which
source code is not available)~\cite{Spjuth2010}.

@@

We have noted that a molecular structure can be characterized using
many numerical descriptors. In this section we describe a
categorization of descriptors. Admittedly, the grouping is somewhat
arbitrary but serves as a useful guide. In addition, the
categorization also takes into account the nature of the chemical
structure being considered - some descriptors are only useful when
applied to small molecules, whereas other descriptors are defined
specifically for polymers or protein structures.  Broadly, one can
classify descriptors based on the nature of structural information
they require.

@@

Topological descriptors take into account connectivity along with atom
and bond labels. They consider the molecule as a labeled graph and
characterize it using graph invariants. In addition to graph
invariants, topological descriptors based on information theory such
as entropy have also been defined\cite{Dehmer:2009uq}. An advantage of
these descriptors is that do not require any intensive pre-processing
steps such as 3D coordinate generation and conformational analysis. In
addition, many graph invariants can be rapidly computed for graphs
corresponding to small molecule structures (even though they may be
computationally intractable on much larger graphs). While this class
of descriptors have been used extensively in predictive modeling
applications
\cite{Garcia-Domenech:2008aa,Randic:2001ad,Besalu:2001aa}, a key
drawback is their lack of interpretability. The abstract nature of
many graph invariants and information theoretic descriptors, their use
in predictive models makes it difficult to explain the model in simple
physicochemical terms, though various attempts have been made at
interpretations \cite{Todeschini:1975dq,Stanton:2003aa}.

Geometric descriptors require a 3D conformation as input and therefore
the computations are not as fast as topological descriptors which do
not require the 3D coordinate generation step. Nearly all geometric
descriptors work on a single conformation and for cases where multiple
conformation must be considered, an averaging procedure is usually
employed. Examples of these descriptors include the gravitational
index \cite{Katritzky:1996ly} and moment of inertia descriptors.
Geometric descriptors encompass many aspects of a molecular
structure. For example, a number of them characterize the molecular
surface. The simplest such example is the surface area descriptor (and
the corresponding volume descriptor). A simple extension of such a
descriptor is to characterize distributions of a physicochemical
property over the molecular surface such as the Charged Partial
Surface Area descriptors \cite{Stanton:1990aa} that characterize the
partial charge distribution over the molecular surfce. Note that
surface descriptor calculations based on analytical
\cite{Connolly:1983aa} or tesselation algorithms
\cite{Eisenhaber:1995qf} can be slow and that parametrized methods
such as the Topological Polar Surface Area (TPSA) \cite{Ertl:2000aa}
and VSA descriptors \cite{Labute:2008aa} that are based on
precalculated surface area values derived from a list of functional
groups.

It is important to note that there is a difference between the
information a descriptor captures and how it is calculated. The above
categories are based on input requirements. This gives rise to
situations where the molecular volume property can be
described by geometrical descriptors, e.g. using an analytical
approaches ~\cite{Eisenhaber:1995qf}, and with approximate,
constitutional descriptors, such as the very fast algorithm use by the
VABC approach~\cite{Zhao:2003fk} which is based on group
contributions.  The latter descriptor has been implemented in the CDK
and correlates well with other approximations to the molecular volume
(Figure \ref{fig:molvol}).

@@

\subsection{Describing other chemistries}

@@

\textbf{Descriptors for polymeric systems} These systems represent a
challenge for descriptor development. There have been reports that
employed traditional small-molecule descriptors
\cite{Mattioni:2002dq,Katritzky:1998cr} that consider the polymer in
terms of the individual monomers. A number of approaches have been
described that includes both deterministic, geometry derived
descriptors \cite{Edvinsson:2003bh} as well as stochastic methods
\cite{Gonzales-Diaz:2003ly}. For the case of polypeptides the
descriptors generally tend to be based on the individual amino
acids. Examples include the VHSE family of descriptors described by
Mei et al \cite{Mei:2005kx}, a set of side chain descriptors from
Collantes and Dunn \cite{Collantes:1995vn}, field based descriptors
\cite{Norinder:1991ys} and the AAindex database
\cite{Kawashima:1999zr} which allows rapid lookup of amino acid
properties.

@@

Other methods developed for crystal structures have been developed
too, for example, using a radial distribution
functions~\cite{Willighagen2005}, which capture the packing of organic
crystal structures. Such approaches have been used to cluster crystal
structures~\cite{Wehrens2006} and compare
properties~\cite{Willighagen2007}.

@@

However, there are certain scenarios where some of the above
characteristics are \emph{undesirable}. One such scenario is in the
safe exchange of structures, such that two parties can perform
analyses on the structure-derived data, without having to know the
structures explicitly. In such cases, the ability to go from a set of
descriptor values back to a chemical structure should be minimized. It
has not been conclusively proven that such a ``one-way'' descriptor
can be defined. For example Masek et al \cite{Masek:2008aa} have shown
that one can regenerate a target structure (or substructural features)
or a set of closely related analogs given a combination of descriptors
such as BCUTs. Tetko \textit{et al.} \cite{Tetko:2005fk} suggest an approach to
prevent reverse engineering but avoid the use of descriptors and
instead sharing structures of related molecules, rather than the
molecules of interest. One could also employ degenerate descriptors,
which would represent a large family of molecules, such that the
target molecule would be hidden. This is likely unreliable as it would
be difficult to perform analyses (e.g., predictive models) with such
descriptors.

@@

As noted above, different implementations of the same descriptor can
generate slightly different values depending on the underlying
chemistry model, differences in reference values and so on. Even
within the same software implementation, different versions of the
same descriptor can generate different values. Thus to maintain
reproducibility in studies that employ molecular descriptors, some
form of versioning is crucial. The simplest form of version is to
manually note the version of the implementation used for the
calculation and report it along with the results. Though useful, it is
not amenable to automation. In some cases, toolkits will provide an
API call to retrieve the version information, which can then be
associated with a descriptor calculation.
